\chapter{Discussion}

\section{Interpretation of Results}
This chapter interprets the experimental findings presented in Chapter 4 and connects them back to the research questions stated in Chapter 1. At the current stage of the project, the strongest evidence comes from the two completed transfer-learning experiments. These results already provide a useful basis for discussing model behavior, the value of pretrained representations, and the role of explainability in this capstone.

Among the completed experiments, ResNet50 with transfer learning achieved the best overall balance of predictive performance. It produced the highest test accuracy, recall, F1-score, and AUC, while DenseNet121 with transfer learning produced slightly lower AUC but higher precision and specificity. This indicates that the two models reached different operating points under the same binary formulation. ResNet50 was better at identifying abnormal images, whereas DenseNet121 behaved more conservatively and produced fewer false alarms. In practical terms, this means that ResNet50 was better aligned with the screening-oriented objective of reducing missed abnormal cases, while DenseNet121 was stronger at avoiding unnecessary positive predictions.

These findings also suggest that transfer learning is an effective strategy for this task. Both pretrained models achieved test AUC values above 0.74, which indicates that ImageNet-pretrained representations transferred well to chest X-ray classification despite the domain difference between natural images and medical images. This supports the argument made in the literature that pretrained convolutional features can provide a useful starting point even when the downstream task involves grayscale radiographs rather than natural color images. In the context of this project, transfer learning was not only practical from a training-efficiency perspective, but also strong enough to produce defensible experimental results for the thesis.

\section{Implications for Medical Image Classification}
The results have several implications for practical chest X-ray classification research at the capstone level. First, they show that a relatively focused binary classification problem can still surface meaningful tradeoffs between model architectures. Even when two models achieve similar AUC values, their recall, precision, and specificity can differ substantially. This is important because medical imaging tasks are rarely judged by accuracy alone. In a clinical screening setting, a model that misses too many abnormal images may be less useful even if its overall accuracy remains acceptable.

Second, the Grad-CAM results provide moderate evidence that the best-performing model learned clinically plausible spatial patterns. In the correctly classified cases, the heatmaps concentrated mainly on thoracic regions instead of obvious background areas, which supports the claim that the model relied on image content relevant to the chest. This improves the interpretability of the result and makes the model behavior easier to explain in the thesis and in the defense. However, the false positive and false negative examples also demonstrate that plausible visual attention does not guarantee correct reasoning. Grad-CAM should therefore be treated as a transparency tool rather than as proof that the model has learned clinically reliable causal structure.

Third, the completed experiments reinforce the value of keeping the project scope controlled. By restricting the task to chest X-ray binary classification with two architectures and one explainability method, the project was able to produce complete training results, interpretable visual evidence, and a clear comparison that can be defended academically. This is preferable to attempting a much broader medical imaging project that would risk incomplete implementation or weak experimental depth.

\section{Limitations}
Several limitations should be acknowledged when interpreting the current results. The first is dataset quality. NIH ChestXray14 is a widely used public benchmark, but its labels were derived automatically from radiology reports rather than assigned manually for this exact classification task. As a result, label noise is likely present. This limits the confidence with which the measured performance can be interpreted as a true estimate of clinical diagnostic quality.

The second limitation is the binary task formulation itself. In this thesis, all non-``No Finding'' labels were grouped into a single abnormal class. This simplification was useful for making the capstone achievable and for supporting a clean comparison across models, but it also removes important clinical detail. Different abnormalities have different visual signatures and different levels of difficulty. A model that performs reasonably on binary normal-versus-abnormal classification may still struggle on more realistic multi-label or disease-specific tasks.

The third limitation is experimental coverage. At present, the strongest conclusions are based on the transfer-learning experiments, while the scratch-training experiments remain to be completed. This means that the project already supports discussion of architecture tradeoffs and explainability, but it does not yet provide the full four-way comparison originally planned. Similarly, only two model architectures were examined, so the conclusions should not be generalized to all deep learning approaches for chest X-ray analysis.

The fourth limitation concerns explainability. Grad-CAM provides an intuitive visualization of the image regions that most influence the prediction, but it remains a qualitative method. Heatmaps can look plausible even when the underlying prediction is wrong, and different explainability methods may highlight different regions. For this reason, the Grad-CAM analysis in this thesis should be interpreted as supportive evidence about model attention rather than as a definitive account of how the model reasons.

\section{Summary}
In summary, the completed experiments indicate that transfer learning is effective for the binary chest X-ray classification task studied in this thesis. ResNet50 with transfer learning currently provides the most balanced overall performance, while DenseNet121 with transfer learning offers higher precision and specificity at the cost of lower recall. The Grad-CAM results further suggest that the best-performing model generally attends to plausible thoracic regions, although its error cases show that explainability does not eliminate uncertainty or guarantee reliable prediction.

These findings support the central argument of the thesis: transfer learning and explainable deep learning can improve the practicality and interpretability of chest X-ray classification in a focused capstone setting. At the same time, the discussion makes clear that the conclusions are bounded by dataset limitations, the simplified binary label design, and the incomplete scratch-model comparison. The next chapter therefore summarizes the main contributions of the project and outlines directions for future improvement.
