\chapter{Experiments and Results}

\section{Experimental Setup}
This chapter presents the experimental results for the four core model configurations defined in Chapter 3. The goal is to compare architecture choice and training strategy under a common evaluation setting. All experiments use the same binary classification task, the same preprocessing pipeline, and the same train-validation-test split. This ensures that any observed performance differences can be interpreted mainly in terms of the model architecture or the use of transfer learning rather than changes in data preparation.

The four experiments evaluated in this thesis are listed below:
\begin{itemize}
    \item E1: ResNet50 trained from scratch;
    \item E2: ResNet50 with transfer learning;
    \item E3: DenseNet121 trained from scratch;
    \item E4: DenseNet121 with transfer learning.
\end{itemize}

The experiments reported in this chapter use the finalized NIH ChestXray14 binary split introduced in Chapter 3, with 78,202 training images, 17,099 validation images, and 16,819 test images. All transfer-learning experiments were trained with an image size of 224, batch size of 16, Adam optimizer, learning rate of 0.001, weight decay of 0.0001, and a maximum of 20 epochs with early stopping patience set to 5. The same loss function, preprocessing pipeline, and evaluation metrics were retained across runs to keep the comparison controlled and interpretable.

\begin{table}[htbp]
    \centering
    \caption{Summary of experimental configurations}
    \label{tab:experiment-setup}
    \begin{tabular}{llll}
        \toprule
        Experiment ID & Model & Training Strategy & Pretrained Weights \\
        \midrule
        E1 & ResNet50 & From scratch & No \\
        E2 & ResNet50 & Transfer learning & ImageNet \\
        E3 & DenseNet121 & From scratch & No \\
        E4 & DenseNet121 & Transfer learning & ImageNet \\
        \bottomrule
    \end{tabular}
\end{table}

\section{Baseline Results}
This section reports the results of the two scratch-training experiments. These baselines are important because they establish how well each model performs without relying on pretrained visual features. They also provide a direct reference point for evaluating the later transfer-learning experiments.

For each model, the final version of this section should report test accuracy, precision, recall, F1-score, AUC, and, if available, sensitivity and specificity. Training time should also be recorded because model efficiency is part of the overall comparison. In addition to the numerical metrics, training and validation loss curves may be included to show convergence behavior. Confusion matrices are useful here because they reveal whether the baseline models struggle more with false positives or false negatives.

\begin{table}[htbp]
    \centering
    \caption{Baseline results for models trained from scratch}
    \label{tab:baseline-results}
    \begin{tabular}{lcccccc}
        \toprule
        Model & Accuracy & Precision & Recall & F1-score & AUC & Training Time \\
        \midrule
        ResNet50 & TBD & TBD & TBD & TBD & TBD & TBD \\
        DenseNet121 & TBD & TBD & TBD & TBD & TBD & TBD \\
        \bottomrule
    \end{tabular}
\end{table}

The final written analysis in this section should answer the following questions:
\begin{itemize}
    \item Which architecture performs better when both are trained from scratch?
    \item Are the scratch-training results already strong enough for the task?
    \item What error patterns are visible in the confusion matrices?
\end{itemize}

\section{Transfer Learning Results}
This section presents the results obtained when ResNet50 and DenseNet121 are initialized with ImageNet-pretrained weights. The purpose is to determine whether transfer learning improves performance relative to scratch training and whether the gains are consistent across both architectures.

Transfer learning was evaluated using the same test metrics for both architectures. In the completed experiments, both pretrained models achieved test AUC values above 0.74, showing that pretrained visual representations transferred effectively to this binary chest X-ray task. However, the two models produced noticeably different error profiles. ResNet50 delivered the strongest overall balance across accuracy, recall, F1-score, and AUC, whereas DenseNet121 achieved higher precision and specificity at the cost of substantially lower recall.

\begin{table}[htbp]
    \centering
    \caption{Results for models trained with transfer learning}
    \label{tab:transfer-results}
    \resizebox{\textwidth}{!}{%
    \begin{tabular}{lcccccc}
        \toprule
        Model & Accuracy & Precision & Recall & F1-score & AUC & Training Time \\
        \midrule
        ResNet50 + transfer learning & 0.6955 & 0.6847 & 0.6272 & 0.6547 & 0.7454 & N/A \\
        DenseNet121 + transfer learning & 0.6711 & 0.7225 & 0.4634 & 0.5646 & 0.7421 & N/A \\
        \bottomrule
    \end{tabular}%
    }
\end{table}

The ResNet50 transfer-learning run completed all 20 epochs and achieved its best validation AUC at epoch 20. On the test set, it reached an AUC of 0.7454, accuracy of 0.6955, and F1-score of 0.6547. These values indicate that the model maintained a relatively balanced tradeoff between sensitivity and specificity. The DenseNet121 transfer-learning run was resumed on a second machine from the checkpoint saved after epoch 7 and stopped early at epoch 17. Its best validation AUC was 0.7348 at epoch 12, and its final test AUC was 0.7421. Although the AUC difference between the two models is small, the DenseNet121 model produced a much lower recall of 0.4634, which indicates that more abnormal cases were missed under the current threshold.

At the same time, DenseNet121 achieved higher precision (0.7225) and specificity (0.8482) than ResNet50, suggesting that it was more conservative in assigning the abnormal label. This tradeoff is important in a medical screening context. A higher-recall model reduces the risk of missed abnormal cases, whereas a higher-specificity model reduces the number of false alarms. Under the binary formulation used in this study, ResNet50 appears to be the better balanced model, while DenseNet121 may be preferred only when the reduction of false positives is the dominant requirement.

\section{Comparative Analysis}
After presenting baseline and transfer-learning results separately, all four experiments should be compared in one integrated analysis. This section is the core of the chapter because it connects the numerical results back to the thesis research questions. The goal is not only to identify the best-performing model, but also to explain the tradeoffs between performance, efficiency, and interpretability.

\begin{table}[htbp]
    \centering
    \caption{Final comparison of all four experiments}
    \label{tab:final-comparison}
    \resizebox{\textwidth}{!}{%
    \begin{tabular}{lccccccc}
        \toprule
        Experiment & Accuracy & Precision & Recall & F1-score & AUC & Sensitivity & Specificity \\
        \midrule
        E1: ResNet50 scratch & TBD & TBD & TBD & TBD & TBD & TBD & TBD \\
        E2: ResNet50 transfer & 0.6955 & 0.6847 & 0.6272 & 0.6547 & 0.7454 & 0.6272 & 0.7537 \\
        E3: DenseNet121 scratch & TBD & TBD & TBD & TBD & TBD & TBD & TBD \\
        E4: DenseNet121 transfer & 0.6711 & 0.7225 & 0.4634 & 0.5646 & 0.7421 & 0.4634 & 0.8482 \\
        \bottomrule
    \end{tabular}%
    }
\end{table}

Among the experiments completed so far, ResNet50 with transfer learning achieved the best overall performance. It produced the highest accuracy, recall, F1-score, and AUC, making it the strongest candidate for the later Grad-CAM analysis. DenseNet121 with transfer learning remained competitive in AUC and clearly outperformed ResNet50 in precision and specificity, but its substantially lower recall means that more abnormal images were classified as normal.

Taken together, the current results suggest that transfer learning is effective for both architectures, but the two models operate at different practical tradeoff points. ResNet50 is more suitable when the goal is a balanced screening-oriented classifier that better preserves abnormal-case detection, while DenseNet121 behaves more conservatively and may be more appropriate when limiting false positives is the stronger priority. The two scratch-training experiments remain to be completed before the full four-way comparison can be finalized.

\section{Training Curves and Error Analysis}
This section should present supporting figures that explain how the models behaved during optimization. The most useful plots are training loss curves, validation loss curves, ROC curves, and confusion matrices. These figures help reveal whether the models converged smoothly, overfit the training data, or struggled with particular classes.

In the final thesis, include representative visualizations such as:
\begin{itemize}
    \item training and validation loss curves for the best baseline model;
    \item training and validation loss curves for the best transfer-learning model;
    \item ROC curves for all four experiments;
    \item confusion matrices for the most important comparisons.
\end{itemize}

The error analysis should focus on whether abnormal images are being missed, whether the model produces too many false alarms, and whether transfer learning changes the balance between these error types. This section will be especially helpful when connecting numerical evaluation to the later Grad-CAM analysis.

\section{Grad-CAM Visualization Results}
The final section of this chapter should present representative Grad-CAM visualizations from the best-performing model and, if time allows, from one weaker comparison model. These visualizations are intended to support interpretability rather than replace quantitative evaluation. For a balanced discussion, include examples from the following groups:
\begin{itemize}
    \item correct normal predictions;
    \item correct abnormal predictions;
    \item false positives;
    \item false negatives.
\end{itemize}

The purpose of the Grad-CAM analysis is to evaluate whether the model attends to clinically plausible image regions. If the highlighted areas correspond to meaningful thoracic regions, this supports the argument that the model has learned useful visual structure. If the model focuses on irrelevant areas such as borders, acquisition artifacts, or background patterns, that behavior should be discussed critically.

In the final version of this section, insert 4 to 8 selected Grad-CAM figures and provide a short interpretation for each one. The analysis should explain what the heatmap suggests about model reasoning and how the visual evidence relates to the model's quantitative strengths and weaknesses.
