\chapter{Literature Review}

This chapter reviews the main research areas that support this thesis: deep learning for chest X-ray analysis, transfer learning in medical image classification, and explainable artificial intelligence for model interpretation. The purpose of the review is not only to summarize prior work, but also to show why the present study is necessary and how its scope is justified. Since this capstone focuses on binary chest X-ray classification using ResNet50, DenseNet121, transfer learning, and Grad-CAM, the literature review is organized around those same themes.

\section{Deep Learning for Chest X-ray Classification}
Chest X-ray classification has become one of the most important applications of deep learning in medical imaging because chest radiography is widely available, relatively inexpensive, and frequently used in routine diagnosis. Unlike some imaging modalities that require expensive acquisition or complex annotation pipelines, chest X-ray datasets can be collected at scale, making them suitable for data-driven classification research. A major milestone in this area was the release of the ChestX-ray8 dataset, later expanded into ChestX-ray14, which provided a hospital-scale benchmark for thoracic disease classification and weakly supervised localization \cite{ref:wang2017}. This dataset played an important role in making chest X-ray classification a standard benchmark problem in medical image analysis.

After the release of large public datasets, many studies applied convolutional neural networks to classify thoracic abnormalities from frontal chest radiographs. These studies differ in task formulation, with some focusing on binary classification and others using multi-class or multi-label settings. In practice, multi-label chest X-ray classification is more clinically realistic because one image may contain multiple findings. However, binary classification remains attractive for focused research because it simplifies the learning objective, reduces ambiguity in label interpretation, and makes controlled model comparison easier. For this reason, binary normal-versus-abnormal classification is a reasonable choice for an undergraduate capstone, especially when the goal is to compare model architectures and training strategies rather than to solve every aspect of the clinical problem at once.

One of the best-known chest X-ray studies is CheXNet, which used a DenseNet121-based model for pneumonia detection and demonstrated that deep neural networks could achieve strong predictive performance on a chest X-ray benchmark \cite{ref:rajpurkar2017}. Later work such as CheXpert extended the discussion by focusing on uncertainty-aware labeling and more systematic benchmark design \cite{ref:irvin2019}. Comparative work by Baltruschat et al. further showed that model architecture choice, additional patient information, and training configuration can all affect chest X-ray classification performance on large public datasets \cite{ref:baltruschat2019}. Together, these studies show that chest X-ray classification is both practically important and technically challenging. They also reveal recurring issues, including label noise, dataset bias, class imbalance, and the gap between benchmark accuracy and trustworthy clinical use.

\section{ResNet and DenseNet in Medical Imaging}
ResNet and DenseNet are among the most influential convolutional neural network architectures used in image classification. ResNet introduced residual learning, which allows deep networks to be optimized more effectively by learning residual mappings rather than attempting to fit every layer from scratch \cite{ref:he2016}. This design helped address optimization difficulties in very deep networks and made residual architectures a standard choice across computer vision tasks. Because of its strong performance and wide adoption, ResNet is often used as a baseline in medical imaging studies.

DenseNet extended this line of development by connecting each layer to all subsequent layers within a dense block, which improves feature reuse and strengthens gradient flow during training \cite{ref:huang2017}. Dense connectivity can be particularly useful when data are limited, because later layers have direct access to earlier representations without requiring repeated relearning of similar low-level features. This property has made DenseNet especially popular in medical image classification, including chest X-ray analysis, where datasets can be large in raw image count but still limited in annotation quality and disease balance.

In chest X-ray research, both ResNet and DenseNet are common because they offer a strong tradeoff between performance, implementation simplicity, and computational feasibility. DenseNet121 became especially prominent through CheXNet \cite{ref:rajpurkar2017}, while ResNet variants remain widely used as reference baselines in medical imaging literature. For this thesis, selecting ResNet50 and DenseNet121 is appropriate for three reasons. First, both are mature and well-understood architectures. Second, both have pretrained ImageNet weights available, making transfer-learning experiments straightforward. Third, the comparison between them is academically meaningful while remaining realistic for a capstone project with limited time and computational resources.

\section{Transfer Learning in Medical Image Analysis}
Transfer learning is widely used in medical image analysis because the amount of labeled medical data is often much smaller than the amount of labeled natural-image data used to train large vision models. A common strategy is to initialize a model with ImageNet-pretrained weights and then fine-tune it on a target medical dataset. This approach can improve convergence speed, stabilize training, and sometimes improve final predictive performance, especially when annotations are limited or class imbalance is severe. Surveys of deep learning in medical image analysis consistently identify transfer learning as one of the most practical techniques for adapting general computer vision models to specialized clinical tasks \cite{ref:litjens2017}.

Prior studies have shown that the benefits of transfer learning are real but not uniform. Pan and Yang provided one of the foundational surveys of transfer learning and clarified why transferring knowledge across domains can reduce the demand for expensive labeled data \cite{ref:pan2010}. Tajbakhsh et al. compared full training and transfer-learning strategies across multiple medical imaging applications and found that transfer learning can be highly effective, particularly when labeled training data are scarce \cite{ref:tajbakhsh2016}. Other studies in radiology and computer-aided detection have also used pretrained CNN backbones to compensate for limited domain-specific supervision \cite{ref:shin2016}. At the same time, the magnitude of transfer-learning gains depends on factors such as the similarity between source and target domains, the depth of fine-tuning, the size of the target dataset, and the quality of labels.

This issue is especially relevant for the present thesis. Although NIH ChestXray14 contains many images, the labels are derived from report mining rather than direct expert annotation for every task, which introduces uncertainty and noise. In addition, some prior studies have explored whether high-capacity models can be trained effectively from scratch by exploiting label dependencies rather than relying entirely on pretraining \cite{ref:yao2017}. Therefore, comparing scratch training with transfer learning is not only a standard engineering choice, but also a meaningful research question. In this capstone, transfer learning is treated as a central variable rather than as an assumed default, because the goal is to measure whether pretrained representations provide a consistent advantage for the selected binary classification task.

\section{Explainable Artificial Intelligence for Chest X-ray Models}
High predictive performance alone is insufficient for safety-critical applications such as medical diagnosis. In healthcare settings, clinicians and researchers need at least some visibility into why a model makes a decision, especially when a prediction may influence follow-up examination or treatment. As a result, explainable artificial intelligence has become an important companion to medical image classification research. Instead of treating the neural network as a complete black box, explainability methods attempt to identify which image regions or features most strongly influence a prediction.

Among these methods, Grad-CAM is one of the most widely used because it is practical, architecture-friendly, and relatively easy to interpret \cite{ref:selvaraju2017}. Grad-CAM computes class-specific gradient information and projects it back onto feature maps in order to highlight image regions associated with the predicted class. In chest X-ray classification, this makes it possible to inspect whether the model is focusing on plausible thoracic regions or on irrelevant artifacts, borders, or text markers. Even though Grad-CAM is not a complete explanation of model reasoning, it provides a useful qualitative tool for detecting suspicious behavior and supporting error analysis.

Chest X-ray studies that include visual explanation often use saliency-based methods to demonstrate that a model is not merely fitting spurious patterns. However, explainability claims in the literature should still be treated carefully. A visually appealing heatmap does not guarantee causal understanding, and interpretation remains partly subjective. Recent surveys on explainable AI in healthcare and medical imaging also emphasize that transparency tools should be treated as supportive evidence rather than as complete validation of model trustworthiness \cite{ref:chaddad2023,ref:bhati2024}. For this thesis, Grad-CAM is an appropriate choice because it adds an interpretable analysis layer without greatly increasing implementation complexity. It fits the capstone scope well by supporting a focused examination of both correct predictions and failure cases.

\section{Research Gap}
The literature reviewed above shows that chest X-ray classification is already a mature research area, but it also reveals several opportunities for a focused study. First, many papers emphasize headline performance on large benchmark tasks, while fewer present a clean and controlled comparison between widely used baseline architectures under both scratch-training and transfer-learning settings. Second, chest X-ray papers often prioritize predictive accuracy, whereas interpretability and error analysis are discussed more briefly or only through a few visual examples. Third, large multi-label formulations can make it difficult to isolate the effect of architecture and training strategy.

This thesis addresses a narrower but useful gap by conducting a compact comparative study centered on one public dataset, one binary task, two standard CNN architectures, and one explainability method. Rather than proposing a new architecture, the study aims to answer a practical question: for binary normal-versus-abnormal chest X-ray classification on NIH ChestXray14, how much do architecture choice, transfer learning, and Grad-CAM-based interpretation contribute to a reliable and understandable pipeline? This narrower framing is suitable for a capstone because it favors clarity, reproducibility, and defensible analysis over unnecessary methodological sprawl.

\section{Summary}
The literature shows that chest X-ray classification is an important and active research area in medical image analysis. Public datasets such as ChestXray14 have enabled large-scale benchmarking, while architectures such as ResNet and DenseNet have become standard choices for visual recognition tasks. Prior work also shows that transfer learning is often beneficial in medical imaging, although its impact depends on the exact dataset and training setup. At the same time, explainability remains essential because strong predictive performance alone is insufficient for trustworthy medical AI.

Based on this review, the methodological choices in this thesis are well justified. ResNet50 and DenseNet121 provide meaningful and established baselines, transfer learning is a central experimental variable, and Grad-CAM offers a practical way to inspect model behavior. The next chapter therefore describes how these components are implemented and evaluated on NIH ChestXray14 for binary chest X-ray classification.
