\chapter{Conclusion and Future Work}

\section{Conclusion and Implications}
This thesis investigated whether transfer learning and explainable deep learning could improve chest X-ray diagnostic classification in a focused capstone setting. The study used the NIH ChestXray14 dataset and formulated the task as binary classification, where images labeled ``No Finding'' were treated as normal and images with any pathology label were treated as abnormal. Two convolutional neural network architectures, ResNet50 and DenseNet121, were selected as the main comparison models because both are well-established in image classification and have been widely used in medical imaging research. Four controlled experiments were completed in total: ResNet50 from scratch, ResNet50 with transfer learning, DenseNet121 from scratch, and DenseNet121 with transfer learning.

The final experimental results show that the best outcome depends on how performance is defined. DenseNet121 trained from scratch achieved the highest test AUC at 0.7457, while ResNet50 with transfer learning delivered the strongest overall balance of accuracy, recall, and F1-score, reaching a test accuracy of 0.6955, F1-score of 0.6547, and AUC of 0.7454. DenseNet121 with transfer learning produced a similar AUC of 0.7421 and the highest precision and specificity, but at the cost of noticeably lower recall. ResNet50 from scratch served as the weakest of the four runs, with a test AUC of 0.7225. These results indicate that the four configurations reached different tradeoff points under the same binary task formulation, and that transfer learning was clearly beneficial for ResNet50 but not uniformly beneficial for DenseNet121.

The Grad-CAM analysis further supported the practical value of the best-performing model. In representative correct predictions, the heatmaps concentrated primarily on thoracic regions rather than on irrelevant image background, suggesting that the model learned clinically plausible visual patterns. At the same time, the false positive and false negative cases showed that plausible attention maps do not guarantee correct decisions. This confirms that explainability should be used to support interpretation of model behavior rather than to replace quantitative evaluation.

Overall, the main implication of this thesis is that a carefully scoped workflow combining controlled experimentation, reproducible evaluation, and explainability can produce academically meaningful results for medical image analysis at the capstone level. Even with a simplified binary problem and only two model architectures, the study was able to reveal clear tradeoffs between sensitivity, specificity, precision, architecture choice, and training strategy. This demonstrates the value of combining model-performance analysis with visual explanation when studying chest X-ray classification.

\section{Future Work}
Several directions can extend this work in a more advanced study. First, the binary classification setting can be expanded into multi-label classification so that individual thoracic diseases are modeled directly instead of being merged into a single abnormal class. This would make the task more clinically realistic and allow model performance to be analyzed at the disease level.

Second, the experimental comparison can be broadened. Additional architectures such as EfficientNet, vision transformers, or chest X-ray models trained with self-supervised learning could be evaluated to determine whether the observed ResNet50--DenseNet121 tradeoff generalizes to more recent model families. Future work should also test threshold optimization and calibration so that the present accuracy-recall-specificity tradeoffs can be tuned more explicitly for screening or decision-support use cases.

Third, explainability can be improved beyond the current Grad-CAM analysis. Future work could compare Grad-CAM with other explainability methods, incorporate uncertainty estimation, or involve expert review of model heatmaps. This would help determine whether the highlighted regions are not only visually plausible, but also clinically meaningful from a radiological perspective.

Finally, the generalizability of the findings should be tested on additional datasets and settings. External validation on benchmarks such as CheXpert or other institutional chest X-ray collections would provide stronger evidence about robustness across domains. If the long-term goal is clinical deployment or decision support, future work should also examine threshold selection, calibration, domain shift, and user-facing evaluation rather than relying on a single fixed test setting.
